---
title: "Supplementary R Code - Data preparation and merging"
author: "Luis E Segura"
date: "2023-05-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Load packages
mypackages <- c("here", "tidyverse", "readstata13")

for(i in mypackages){
  library(i, character.only = T)
}
```

Raw data are available in the course Dropbox folder. You will need to update the filepath below to the appropriate location on your computer. Make sure the folder with the datafiles (I named it data) is saved inside your working directory. To view your working directory run the following command getwd(). To set a new working directory use setwd("FILEPATH") and use the FILEPATH for that directory, for example setwd("C:/Luis/Analysis of Complex Survey/").

The original datafiles are in STATA format (.dat). We will use the package readstata13 to read them into R.

```{r , }
### assign the names of the Stata datasets to a vector
nsduh_data_names <- list.files(path = here("data/"), pattern = "*.dta") 

### create an empty list where we will store all the datasets as we read them in
nsduh_list <- list()

### Create a vector of the variable names we want to keep
var_names <- c("caseid", "realid", "vestr", "verep", "year", "heryr", "anlyr", "mrjyr", "newrace2", 
               "iranlfy", "age2", "income", "irsex", "pden", "abodanl", "abodher", "anydrug", "analwt_c")

### loop to import the 8 NSDUH datasets into the list we created ("nsduh_list")
for(i in seq_along(nsduh_data_names)){
  
  # Read each dataset into an element of the list nsduh_list
  nsduh_list[[nsduh_data_names[i]]] <- read.dta13(here(paste0("data/", nsduh_data_names[i])),
                                                  convert.factors = T,
                                                  generate.factors = T,
                                                  nonint.factors = T) |> 
    # Convert variable names to lowercase
    rename_all(tolower) |>
    # Create year variable
    mutate(year = if_else(i < 5, 2001 + i, 2003 + i), 
           realid = year * 1e5 + caseid) |>
    # Restrict datasets to just the variables we need for the analysis
    select(var_names)
  
}

### coerce list into a dataframe
nsduh <- nsduh_list |>
  bind_rows()

### Save data

save(nsduh, file = here("data/", "nsduh.RData"))
save(nsduh, file = here("data/", "nsduh.csv"))
```